Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-10 text/manuscript-full-KS.qmd
copy edits

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metropolitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board underwrote the project -- providing data and funding -- and published the book as part of their nascent efforts "to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. Conceptualized as an effort to to "capture and quantify" (p. 11) the knowledge of admissions officers, @RN4982 developed the Market Segment Model by analyzing the SAT score-sending behavior of high school seniors in 1980. The thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets. The left panel of @fig-chicago-recruiting shows Geomarkets in the Chicago-land area. As a planning tool for colleges, the Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

@@

This manuscript analyzes College Board Geomarkets as a case study of quantification. In particular, we draw from the discussions of correlation and homophily from @RN4975. @RN4982 identified the correlates of 1980 SAT score-sending behavior and concluded that student demand for higher education is primarily a function of social class. The Market Segment Model argues that homophily -- actors that share characteristics form connections -- is the organizing principle of competition and college choice, stating that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Scholarship on quantification demonstrates that making recommendations based on past correlations amplifies the effects of historic structural inequality [@RN4794]. The snapshot of student demand in 1980 -- itself a consequence historic structural inequality -- was programmed into recruiting products that colleges utilize to identify and target prospective students. The result is a supply-side that amplifies structural inequalities observed on the demand-side.

@@

Finally, we discuss implications for scholarship. Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [@vista_eab_ml]. We propose two fruitful approaches to studying the role of third-party vendors in enrollment management. First, case studies -- like this manuscript and @RN5039 -- utilize primary data collection to conduct granular analyses of particular products or vendors. Second, a growing number of subscription databases (e.g., Orbis, Pitchbook, LSEG) serve the information needs of the investment community by providing data on private firms and acquisitions or investments in private firms. These data sources may provide the basis for a scalable "knowledge infrastructure" [@RN5025] that investigates the role of third-party vendors in education.

@@

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. In turn, recruiting visits to high schools -- and subsequent phone calls -- are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly white public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

@@

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

@@

While the contemporary EM industry is characterized by software-as-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The College Board Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The College Board Enrollment Planning Service (EPS), launched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"

@@

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science, in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). In *The company We Keep: Colleges and Their Competition*, @RN4982 [chapter 4] conduct a network analysis whereby two colleges are defined as competitors if a large number of students submit SAT scores to both colleges. @RN4975 observes that network science models often draw from rational choice theory, which leads to the assumption that homophily is the result of voluntary action by individuals.

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@Chun2018Queerying, p. 66]. We observe similar phenomena in our case. Based on a network analysis of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side of college choice-- which is a function of historic structural inequality –- is programmed into the supply-side.

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

We make targeted contributions to scholarship on quantification in the sociology of education. As @RN3519 demonstrates in the chapter *Numbers*, EM is fundamentally concerned with quantification. However, sociological analyses of EM focus on the behavior of colleges and their agents, creating the sense that inequalities created by EM are a function of individual colleges reacting to macro structures in their environment such as rankings and financial pressures [e.g., @RN5003; @RN3519]. To our knowledge, the sociology of education has not considered producer-facing products that quantify students on behalf of colleges. Market research and investigative journalism indicates that the vast majority of colleges purchase third-party EM products to identify prospects and to decide which schools and communities to target [@RN5027]. Recruiting products grounded in the logic of predictive analytics take a snapshot of student demand -- without considering historical structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches vertically categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Third-party EM products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociology.

```{=html}
<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->
```
This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advises colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

**Correlations**. @RN4982 [Chapter 3], *A Sense of Place: Students, Families, and Communities*, investigates the student characteristics correlated with being in the local, in-state, regional, or national market segment. The analyses identify four variables -- educational aspirations, parental education, scholastic aptitude, and family income -- that predict score sending behavior, both individually and in combination. Because these four variables "reflect the basic social patterns of the nation, it would have been surprising if these were not the four social variables that best explained the patterns of college choice" [@RN4982, p. 33].<!--
and "The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice" (pp. 34-35). \[QUOTES REDUNDANT\]
-->

@@

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS), based on the Market Segment Model [@RN4987; @RN4980]. EPS was an early software-as-service platform that recreated the analyses of @RN4982. For each Geomarket, colleges could obtain Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors. Based on background conversations with EM professionals, EPS software also provided information about the score-sending behavior of individual high schools within each Geomarket. Therefore, colleges used EPS software to decide which Geomarkets to recruit from and which high schools to visit within targeted Geomarkets. Typical @RN5019 marketing material describes EPS as,

@@

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is a social science depiction. However, the College Board inscribed Geomarkets and the Market Segment Model into EPS software that tells colleges which Geomarkets and high schools to target. Technologies that target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong academic achievement.

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527 \[p. 148\] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities. College access scholarship from economics often analyzes the effects of district, state, and federal policies [e.g., @RN5043]. These literatures ignore the role of third-party vendors, which are often owned by private equity and offer software-as-service platforms that perform core functions for schools and colleges.[^21]

@@

EAB exemplifies the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. By 1995 EM became their primary focus. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

@@

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50 per name). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board's *Student Search Service,* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and whether students used Naviance to "research competitor institutions" [@markup_naviance]. A University of Utah procurement document states the Intersect subscription is necessary because "there is unique group of prospective students who are only in the PowerSchool Naviance platform" [@utah_procurement].

>>>>

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, emails, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metropolitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board underwrote the project -- providing data and funding -- and published the book as part of their nascent efforts "to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. Conceptualized as an effort to "capture and quantify" (p. 11) the knowledge of admissions officers, @RN4982 developed the Market Segment Model by analyzing the SAT score-sending behavior of high school seniors in 1980. The thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets. The left panel of @fig-chicago-recruiting shows Geomarkets in the Chicago-land area. As a planning tool for colleges, the Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

@@

This manuscript analyzes College Board Geomarkets as a case study of quantification. In particular, we draw from the discussions of correlation and homophily from @RN4975. @RN4982 identified the correlates of 1980 SAT score-sending behavior and concluded that student demand for higher education is primarily a function of social class. The Market Segment Model argues that homophily -- actors that share characteristics form connections -- is the organizing principle of competition and college choice, stating that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Scholarship on quantification demonstrates that making recommendations based on past correlations amplifies the effects of historic structural inequality [@RN4794]. The snapshot of student demand in 1980 -- itself a consequence of historic structural inequality -- was programmed into recruiting products that colleges utilize to identify and target prospective students. The result is a supply-side that amplifies structural inequalities observed on the demand-side.

@@

Finally, we discuss implications for scholarship. Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private-equity owned firms selling software-as-a-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [@vista_eab_ml]. We propose two fruitful approaches to studying the role of third-party vendors in enrollment management. First, case studies -- like this manuscript and @RN5039 -- utilize primary data collection to conduct granular analyses of particular products or vendors. Second, a growing number of subscription databases (e.g., Orbis, Pitchbook, LSEG) serve the information needs of the investment community by providing data on private firms and acquisitions or investments in private firms. These data sources may provide the basis for a scalable "knowledge infrastructure" [@RN5025] that investigates the role of third-party vendors in education.

@@

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. In turn, recruiting visits to high schools -- and subsequent phone calls -- are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly White public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

@@

Drawing from scholarship on organizational theory, EM processes require colleges to make many "make or buy" [@RN4738; @RN980] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. The most widely-known class of EM vendor is EM consulting firms, which provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success [@RN5027]. As "creating a class" has become complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].[^2]

@@

While the contemporary EM industry is characterized by software-as-a-service products sold to colleges by private equity backed firms [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. Every year, colleges must figure out which prospects should receive promotional material and which high schools to visit. The College Board Student Search Service -- copied by ACT's Educational Opportunity Service -- became the ubiquitous means of converting prospects into leads that could be served targeted recruiting interventions. The College Board Enrollment Planning Service (EPS), launched in 1984, is an early software-as-a-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification that sort students into colleges. Student Search Service includes or excludes prospects from student list purchases by colleges based on the value of selected filters like Geomarket while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. The multi-disciplinary literature on quantification helps conceptualize how third-party products utilize Geomarkets and the potential consequences of this utilization on the inequitable sorting of students into colleges.

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who developed a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared [\ldots], big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"

@@

**Homophily**. Homophily is the idea that actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science, in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). In *The Company We Keep: Colleges and Their Competition*, @RN4982 [chapter 4] conduct a network analysis whereby two colleges are defined as competitors if a large number of students submit SAT scores to both colleges. @RN4975 observes that network science models often draw from rational choice theory, which leads to the assumption that homophily is the result of voluntary action by individuals.

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@Chun2018Queerying, p. 66]. We observe similar phenomena in our case. Based on a network analysis of 1980 SAT score-sending, @RN4982 conclude that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reason that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side of college choice -- which is a function of historic structural inequality -- is programmed into the supply-side.

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that target different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorize students into four market segments -- local, in-state, regional, and national -- based on their score-sending behavior and then evaluate the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

We make targeted contributions to scholarship on quantification in the sociology of education. As @RN3519 demonstrates in the chapter *Numbers*, EM is fundamentally concerned with quantification. However, sociological analyses of EM focus on the behavior of colleges and their agents, creating the sense that inequalities created by EM are a function of individual colleges reacting to macro structures in their environment such as rankings and financial pressures [e.g., @RN5003; @RN3519]. To our knowledge, the sociology of education has not considered producer-facing products that quantify students on behalf of colleges. Market research and investigative journalism indicates that the vast majority of colleges purchase third-party EM products to identify prospects and to decide which schools and communities to target [@RN5027]. Recruiting products grounded in the logic of predictive analytics take a snapshot of student demand -- without considering historical structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches vertically categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. Third-party EM products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-a-service platforms, but have remained invisible to sociology.

```{=html}
<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->
```
This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board Enrollment Planning Service (EPS), which advises colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

**Correlations**. @RN4982 [Chapter 3], *A Sense of Place: Students, Families, and Communities*, investigates the student characteristics correlated with being in the local, in-state, regional, or national market segment. The analyses identify four variables -- educational aspirations, parental education, scholastic aptitude, and family income -- that predict score-sending behavior, both individually and in combination. Because these four variables "reflect the basic social patterns of the nation, it would have been surprising if these were not the four social variables that best explained the patterns of college choice" [@RN4982, p. 33].<!--
and "The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice" (pp. 34-35). \[QUOTES REDUNDANT\]
-->

@@

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS), based on the Market Segment Model [@RN4987; @RN4980]. EPS was an early software-as-a-service platform that recreated the analyses of @RN4982. For each Geomarket, colleges could obtain Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors. Based on background conversations with EM professionals, EPS software also provided information about the score-sending behavior of individual high schools within each Geomarket. Therefore, colleges used EPS software to decide which Geomarkets to recruit from and which high schools to visit within targeted Geomarkets. Typical @RN5019 marketing material describes EPS as,

@@

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is a social science depiction. However, the College Board inscribed Geomarkets and the Market Segment Model into EPS software that tells colleges which Geomarkets and high schools to target. Technologies that target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-White Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-White students with strong academic achievement.

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527 \[p. 148\] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities. College access scholarship from economics often analyzes the effects of district, state, and federal policies [e.g., @RN5043]. These literatures ignore the role of third-party vendors, which are often owned by private equity and offer software-as-a-service platforms that perform core functions for schools and colleges.[^21]

@@

EAB exemplifies the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-a-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. By 1995 EM became their primary focus. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

@@

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50 per name). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-a-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board's *Student Search Service,* contemporary software-as-a-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and whether students used Naviance to "research competitor institutions" [@markup_naviance]. A University of Utah procurement document states the Intersect subscription is necessary because "there is unique group of prospective students who are only in the PowerSchool Naviance platform" [@utah_procurement].
